---
title: "Programming Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mean total number of steps taken per day

First let us read the data from the disk and see what variables it contains.

```{r read_csv, ECHO=TRUE}
activity_data <- read.csv("./data/activity.csv")
str(activity_data)
```

We can see that there are some missing values in the steps variable and the date variable is considered a factor variable and not a date variable. Hence some preprocessing of the data is to be done before any kind of computation can be done.

The missing values are going to be filtered out, and the date are going to be cast to date.

```{r preprocess_data, ECHO=TRUE}
activity_data$date <- as.Date(activity_data$date)
activity_data_noNas <- na.omit(activity_data)
str(activity_data_noNas)
```

Let us now have a quick look at the data statistics.

```{r activity_data, ECHO=TRUE}
summary(activity_data_noNas)
```

Now we are going to compute the total number of steps taken each day by the individual. And let us have a rough look at the data we obtained.

```{r aggregate_data, ECHO=TRUE}
data <- subset(activity_data_noNas, select = -date)
dates <- subset(activity_data_noNas, select = date)
stepsPerDay <- apply(data, 2, function(x) tapply(x, dates, sum))
head(stepsPerDay)
```

Let us get some statistics about the total number of steps per day:

```{r summary_activity, ECHO=TRUE}
summary(stepsPerDay[,1])
```

Below are the mean and the median of this aggreagated data.
```{r mean_of_totals, ECHO=TRUE}
mean(stepsPerDay[,1])
```
```{r median_of_totals, ECHO=TRUE}
median(stepsPerDay[,1])
```

A picture is worth a million words thus let us create an Histogram of the steps per day variable.  
As to know what is the difference between a bar plot and a histogram, in bar plot columns are positioned over a label that represents a categorical variable whereas in a histogram columns are positioned over a label that represents a continuous, quantitative variable.

```{r histogram_stepsperday, ECHO=TRUE}
library(ggplot2)
stepsPerDayDf <- as.data.frame(stepsPerDay)
maxStepPerDay <- max(stepsPerDayDf$steps)
g <- ggplot(stepsPerDayDf, aes(steps)) +
  geom_histogram(binwidth = ceiling(maxStepPerDay / 5))
g
```

## Average daily activity pattern

Let us compute the average number of steps taken during each interval accross all days.

```{r steps_per_interval, echo=TRUE}
steps <- subset(activity_data_noNas,select=steps)
stepsPerIntervalL <- apply(steps, 2, function(x) tapply(x, activity_data_noNas$interval, mean))
stepsPerIntervalDf <- as.data.frame(stepsPerIntervalL)
rownames(stepsPerIntervalDf) <- NULL
stepsPerIntervalDf$interval <- as.integer(rownames(stepsPerIntervalL))

str(stepsPerIntervalDf)
```

Calling summary on the stepsPerIntervalDf data frame we can see that the average maximum number of steps taken per interval is 206.170 steps.
```{r steps_per_interval_summary, echo=TRUE}
summary(stepsPerIntervalDf)
```

Let us now draw the time series of the average number of steps taken during intervals accross all days
```{r time_serie, echo=TRUE}
grahDf <- data.frame(stepsPerIntervalDf)
max_interval <- stepsPerIntervalDf[which.max(stepsPerIntervalDf$steps),]$interval
grahDf$highlight <- ifelse(grahDf$interval == max_interval, "highlight", "normal")
mycolours <- c("highlight" = "red", "normal" = "black")

ggplot(grahDf, aes(interval, steps)) +
  geom_point(size = 1, aes(colour = highlight)) +
  scale_color_manual("Status", values = mycolours) +
  theme(legend.position = "none") +
  labs(title= "Average # steps taken per interval",
       x = "Interval",
       y = "Average number of steps") +
  geom_line()
```


```{r max_step_interval, echo=TRUE}
stepsPerIntervalDf[which.max(stepsPerIntervalDf$steps),]
```
Thus we can see that on average the maximum number of steps on a 5 minute interval is performed around 8:35 AM.

## Inputing missing values

To get the number of NAs in the data set we can use the summary function.
```{r  activity_data_summary, echo=TRUE}
summary(activity_data)
```

Or we can use the is.na and sum routines to get them manually.
```{r activity_data_nas, echo=TRUE}
sum(is.na(activity_data))
```
Hence, we can see that there are 2304 NA values in the data set. By calling:
```{r activity_data_nas_step, echo=TRUE}
sum(is.na(activity_data$steps))
```
We can see that there are all located in the step variable.

### First option for NA filling

In this first option, I am filling the missing values by computing the mean of the five values preceding the missing value.
```{r fill_NA_gaps, echo=TRUE}
activity_data_NA_filled <- as.data.frame(activity_data)
activity_data_NA_steps <- c(activity_data$steps)
for(i in seq_along(activity_data_NA_steps)) {
  y <- c(0, activity_data_NA_steps)
  if(is.na(activity_data_NA_steps[i])) {
    idx <- seq(i-5,i-1)
    idx[idx < 1] <- 1
    activity_data_NA_steps[i] <- round(mean(y[idx]), digits = 0) 
  }
}
activity_data_NA_filled$steps <- activity_data_NA_steps
```

Now let us have a look at a summary of the data with the NAs filled.
```{r summary_fill_NA_gaps, echo=TRUE}
summary(activity_data_NA_filled)
```

Now let`s calculate the the sum of steps taken per interval across all days.
```{r steps_per_day_filled, echo=TRUE}
data <- subset(activity_data_NA_filled, select = -date)
dates <- subset(activity_data_NA_filled, select = date)
stepsPerDayFilled <- apply(data, 2, function(x) tapply(x, dates, sum))
head(stepsPerDayFilled)
```

Below are the mean and the median of this aggreagated data
```{r mean_of_totals_filled, ECHO=TRUE}
mean(stepsPerDayFilled[,1])
```
```{r median_of_totals_filled, ECHO=TRUE}
median(stepsPerDayFilled[,1])
```

We can see that this method changed the mean and median a lot compared to the data frame with NA omitted.

When taking a look at the histogram of the number of steps, we can see that a lot of 0s were added.
```{r hist_filled_NA_per_day, echo=TRUE}
stepsPerDayFilledDf <- as.data.frame(stepsPerDayFilled)
maxStepPerDay <- max(stepsPerDayFilledDf$steps)
g <- ggplot(stepsPerDayFilledDf, aes(steps)) +
  geom_histogram(binwidth = ceiling(maxStepPerDay / 5))
g
```

Those zeros add bias to the data. We want to avoid that. Let us have a look at the distribution of the NA values in the data.

```{r NA_per_day, echo=TRUE}
steps <- subset(activity_data, select = -date)
dates <- subset(activity_data, select = date)
NAperDay <- apply(steps, 2, function(x) tapply(x, dates, function(y) {sum(is.na(y))}))

NAperDay[NAperDay[,'steps'] > 0,]
```

We can see that we are actually missing only whole days. So let us put in place a better stategy to fill the gaps.

### Second option for NA filling

In this second option, let us compute the mean of steps per interval taken across all days. And filled the missing days with these averages.

Let's compute the mean of steps per interval across all day using apply.
```{r mean_step_per_interval, echo = TRUE}
stepsNoNas <- subset(activity_data_noNas, select=steps)
intervals <- subset(activity_data_noNas, select=interval)
meanStepPerInterval <- apply(stepsNoNas, 2, function(x) tapply(x, intervals, function(y) {round(mean(y), digits=0)}))
```

Now one can fill the missing steps with the vector computed above and by retrieving the dates where the NAs are.
```{r second_NA_filling_option, echo=TRUE}
NADates <- rownames(NAperDay[NAperDay[,"steps"] == 288,])
for(d in NADates) {
     activity_data_NA_filled[activity_data_NA_filled$date == d,"steps"] <- c(meanStepPerInterval)
 }
```

Now let`s calculate the the sum of steps taken per interval across all days.
```{r step_per_day_filled_NA_per_day_second_option, echo=TRUE}
data <- subset(activity_data_NA_filled, select = -date)
dates <- subset(activity_data_NA_filled, select = date)
stepsPerDayFilled <- apply(data, 2, function(x) tapply(x, dates, sum))
```

We can compute the mean and median of this aggreagated data.
```{r mean_of_totals_filled_second_option, ECHO=TRUE}
mean(stepsPerDayFilled[,1])
```
```{r median_of_totals_filled_second_option, ECHO=TRUE}
median(stepsPerDayFilled[,1])
```
The results are now closer to the results we had with the data where NAs where omitted.

Let us have a visualisation of the NA filling
```{r hist_filled_NA_per_day_second_option, echo=TRUE}
stepsPerDayFilledDf <- as.data.frame(stepsPerDayFilled)
maxStepPerDay <- max(stepsPerDayFilledDf$steps)
g <- ggplot(stepsPerDayFilledDf, aes(steps)) +
  geom_histogram(binwidth = ceiling(maxStepPerDay / 5))
g
```
Again the shape of this histogram is much closer to the one when we simply omitted the NAs.

## Differences in activity pattern between Weekdays and Weekend

The code below adds a column with a weekday factor with two levels:
  - weekday
  - weekend
  
```{r weekday_factor, ECHO=TRUE}
activity_data_NA_filled$weekday <- factor(weekdays(activity_data_NA_filled$date))
levels(activity_data_NA_filled$weekday) <- c("weekday", "weekday", "weekday", "weekday", "weekday", "weekend", "weekend")
```

I used the *dplyr* library to split the data frame by interval and weekday factor.
```{r data_prep_for_graph, ECHO=TRUE, message=FALSE}
library(dplyr)
stepsPerIntervalPerWeekdayDf <- activity_data_NA_filled %>% group_by(interval, weekday) %>% summarise(mean_steps = mean(steps))

summary(stepsPerIntervalPerWeekdayDf)
```

Finally, let us draw the graphs with 2 facets that shows the pattern of activity across all days depending on their type.
```{r time_serie_per_weekday, ECHO=TRUE}
ggplot(stepsPerIntervalPerWeekdayDf, aes(interval, mean_steps)) +
  facet_grid(weekday ~ .) +
  theme(legend.position = "none") +
  labs(title= "Average # steps taken per interval",
       x = "Interval",
       y = "Average number of steps") +
  geom_line()
```

